---
title: "Assignment 3"
format: html
editor: visual
author: "Sofia Quijada"
---

```{r}
library(KMsurv)
library(tidyverse)
library(survival)
```

## Question 4.1

**Effect of ploidy on survival of patients with cancer of the tongue. Using data on ANEUPLOID TUMORS only.**

**a) Estimate survival function at 12 months and 60 months after transplant using table 1.6 data. Get SE of estimates**

type: Tumor DNA profile (1=Aneuploid Tumor, 2=Diploid Tumor) time: Time to death or on-study time, weeks delta: Death indicator (0=alive, 1=dead). 0 = censored in this dataset, as per table 1.6 description.

```{r}
data(tongue)

aneuploid <- subset(tongue, type == 1)

#number of people at risk at beginning of study = number of patients we have data for
N <- nrow(aneuploid)
```

Recall our time is in weeks, change at the end.

$\hat{S}(t)$ = 1 if t less than $t_1$ of interest.

$\hat{S}(t)=∏_{ti≤t}(1−d_i/Y_i)$ if t beyond $t_1$ of interest.

```{r}
#order event times, where delta = 1
event_times <- sort(unique(aneuploid$time[aneuploid$delta == 1]))

#start constructing table
km_table <- data.frame(
  time = event_times
)

#compute di, number of deaths at time ti
km_table$d_i <- sapply(
  km_table$time,
  function(t) sum(aneuploid$time == t & aneuploid$delta == 1) #obs with event (delta = 1) at time t
)

#compute ni, number at risk at time ti
km_table$n_i <- sapply(
  km_table$time,
  function(t) sum(aneuploid$time >= t) #obs before time t
) 
```

Compute $\hat{S}(t)=∏_{ti≤t}(1−d_i/n_i)$ for each row. Multiply them to get $\hat{S}(t)$

```{r}
#survival step terms
km_table$surv_step <- 1 - km_table$d_i / km_table$n_i

#multiply to get estimator for S(t) at each t
km_table$S_t <- cumprod(km_table$surv_step)
```

Compute variance and SE for each estimate.

$\widehat{\mathrm{Var}}(\hat{S}(t)) = \hat{S}(t)^2 \sum_{t_i \le t} \frac{d_i}{Y_i (Y_i - d_i)}$

```{r}
#compute sum term 
km_table$greenwood_term <-
  km_table$d_i / (km_table$n_i * (km_table$n_i - km_table$d_i))

#compute cumulative sum
km_table$cum_greenwood <-
  cumsum(km_table$greenwood_term)

#compute greenwood variance by mult cumsum with S(t) estimate for each t
km_table$var_S_t <-
  (km_table$S_t^2) * km_table$cum_greenwood

#compute SE
km_table$SE_S_t <-
  sqrt(km_table$var_S_t)
```

KM estimator uses the last event time ≤ target time.

```{r}
get_survival <- function(target_time) {

  eligible <- which(km_table$time <= target_time)

  if (length(eligible) == 0) {

    return(data.frame(
      time = target_time,
      Survival = 1,
      SE = 0
    ))

  }

  idx <- max(eligible)

  data.frame(
    time = target_time,
    Survival = km_table$S_t[idx],
    SE = km_table$SE_S_t[idx]
  )
}
#conversion factor from months to weeks is 4.345
#input should be in weeks, 12 or 60 months in weeks
S_12 <- get_survival(12*4.345)
S_60 <- get_survival(60*4.345)

S_12
S_60
```

S(t = 12 months) = 0.65, SE = 0.07

S(t = 60 months) = 0.23, SE = 0.09

**b) Estimate the cumulative hazard rate H(t) at 60 months. Find the standard error of** $\hat{H}(t)$ **Estimate S(60) by exp(-** $\hat{H}(t)$**) and compare to part a)**

$\hat{H}(t) = -ln[\hat{S}(t)]$ from KM estimator

```{r}
H_60 <- -log(S_60$Survival)
H_60
```

Use Nelson Aalen cumulative hazard rate estimator instead.

$\hat{H}(t) = \sum_{t_i \le t} \frac{d_i}{n_i}$, or 0 if $t_1$ of interest bigger than t.

$\mathrm{Var}\!\left[\hat{H}(t)\right] = \sum_{t_i \le t} \frac{d_i}{Y_i^2}$

```{r}
#add nelson-aalen increments
km_table$haz_increment <- km_table$d_i / km_table$n_i

#compute cumulative hazard
km_table$H_t <- cumsum(km_table$haz_increment)

#calculate variance and SE
km_table$var_increment <- km_table$d_i / (km_table$n_i^2)

km_table$var_H_t <- cumsum(km_table$var_increment)

km_table$SE_H_t <- sqrt(km_table$var_H_t)

#estimate at 60 months
idx60 <- max(which(km_table$time <= 60*4.345))
H60 <- km_table$H_t[idx60]
SE_H60 <- km_table$SE_H_t[idx60]
H60
SE_H60
```

$\hat{H}(t = 60)$ = 1.39, SE = 0.375.

Estimate S(60) from H: $\hat{S}(t) = \exp\!\left[-\hat{H}(t)\right]$, compare to S(60)

```{r}
S60_from_NA <- exp(-H60)

S_60$Survival #KM
S60_from_NA 
```

The Nelson-Aalen survival estimator derived from the cumulative hazard rate estimator has a slightly higher value.

**c) Find a 95% linear CI for S(60)**

$\hat{S}(t) \pm z_{0.975}\,\mathrm{SE}\!\left(\hat{S}(t)\right)$

```{r}
z <- 1.96

lower_linear <- S_60$Survival - z * S_60$SE
upper_linear <- S_60$Survival + z * S_60$SE

c(lower_linear, upper_linear)
```

\[0.042, 0.42\]. This is a big confidence interval.

**d) Find a 95% log-transformed CI for S(60)**

Confidence interval =$[\hat{S}(t_1)^{1/\theta}, \hat{S}(t_1)^{ \theta}]$,

Where $\theta = exp({z_{0.975}SE(S(t_1))/ln(S(t_1)})$

```{r}
theta <- exp(z * S_60$SE / log(S_60$Survival))
lower_log <- S_60$Survival^(1/theta)
upper_log <- S_60$Survival^theta
c(lower_log, upper_log)
```

\[0.187, 0.273\] 95% CI for S(60) = 0.23, much more reasonable

**e) Find a 95% arcsine-square root CI for S(60)**

Not computing this manually. Use "survival" package.

```{r}
fit <- survfit(Surv(time, delta) ~ 1, data = aneuploid, conf.type = "arcsin")
summary(fit, times = c(12*4.345, 60*4.345))
```

\[0.0744, 0.436\] is 95% arcsin square root survival CI.

**f) Using the log transformation find a 95% EP confidence band for the survival function over the range 3-6 yrs (36-72 months)**

range: 36 \< t \< 72 months

n = sample size

$a_L = \frac{n Var(t_L),}{1+n Var(t_L))}$ , $a_U = \frac{n Var(t_U),}{1+n Var(t_U))}$

Confidence band =$[\hat{S}(t_1)^{1/\theta}, \hat{S}(t_1)^{ \theta}]$,

Where $\theta =exp(\frac{c_\alpha(a_L, a_U) SE(S(t)))}{ln[\hat{S}(t)]})$

$c_\alpha$ found from table C.3 of textbook using values of $a_L$ and $a_u$

```{r}
# bounds in weeks
lower_t <- 36 * 4.345
upper_t <- 72 * 4.345

lower_S <- get_survival(lower_t)
upper_S <- get_survival(upper_t)

# variance = SE^2
a_L <- N * (lower_S$SE)^2 / (1 + N * (lower_S$SE)^2) #0.234
a_U <- N * (upper_S$SE)^2 / (1 + N * (upper_S$SE)^2) #0.321

# from Table C.3 - between 3.06 and 3.02 
c_alpha <- 3.04

theta_lower <- exp(
  c_alpha * lower_S$SE / log(lower_S$Survival)
)

theta_upper <- exp(
  c_alpha * upper_S$SE / log(upper_S$Survival)
)

# correct ordering
lower_band_36 <- lower_S$Survival^(1/theta_lower)
upper_band_36 <- lower_S$Survival^(theta_lower)

lower_band_72 <- upper_S$Survival^(1/theta_upper)
upper_band_72 <- upper_S$Survival^(theta_upper)

print(c(lower_band_36, upper_band_36))

print(c(lower_band_72, upper_band_72))
```

This is just for the end points, here is for the entire time period:

```{r}
# bounds in weeks
lower_t <- 36 * 4.345
upper_t <- 72 * 4.345

lower_S <- get_survival(lower_t)
upper_S <- get_survival(upper_t)

# variance = SE^2
a_L <- N * (lower_S$SE)^2 / (1 + N * (lower_S$SE)^2) #0.234
a_U <- N * (upper_S$SE)^2 / (1 + N * (upper_S$SE)^2) #0.321

# from Table C.3 - between 3.06 and 3.02 
c_alpha <- 3.04

times_weeks <- km_table$time[
  km_table$time >= lower_t &
  km_table$time <= upper_t
]

#compute EP band
EP_band <- lapply(times_weeks, function(t) {

  S <- get_survival(t)

  if (S$Survival == 0) {
    return(data.frame(
      time = t,
      Survival = 0,
      lower = 0,
      upper = 0
    ))
  }

  theta <- exp(
    c_alpha * S$SE / log(S$Survival)
  )

  lower <- S$Survival^(1/theta)
  upper <- S$Survival^(theta)

  data.frame(
    time = t,
    Survival = S$Survival,
    lower = lower,
    upper = upper
  )
})

EP_band <- do.call(rbind, EP_band)
EP_band
```

Since data given is in weeks, there are little observations that go beyond 36 months. I suspect the questions were designed thinking the "time" data was in weeks.

**g) Using log transformation find 95% Hall-Wellner confidence band for the survival function over the range 3-6 yrs (36-72 months)**

$a_L = \frac{n Var(t_L),}{1+n Var(t_L))}$ , $a_U = \frac{n Var(t_U),}{1+n Var(t_U))}$

Confidence band =$[\hat{S}(t_1)^{1/\theta}, \hat{S}(t_1)^{ \theta}]$,

Where $\theta =exp(\frac{k_\alpha(a_L, a_U)(1 + nVar(S(t))}{n^{1/2}ln[\hat{S}(t)]})$

$k_\alpha$ from table C.4 of Appendix C.

```{r}
# bounds in weeks
lower_t <- 36 * 4.345
upper_t <- 72 * 4.345

lower_S <- get_survival(lower_t)
upper_S <- get_survival(upper_t)

# variance = SE^2
a_L <- N * (lower_S$SE)^2 / (1 + N * (lower_S$SE)^2) #0.234
a_U <- N * (upper_S$SE)^2 / (1 + N * (upper_S$SE)^2) #0.321

# from Table C.3 - between 3.06 and 3.02 
k_alpha <- 1 #doesn't seem to exist :c, put in 1 as a placeholder

times_weeks <- km_table$time[
  km_table$time >= lower_t &
  km_table$time <= upper_t
]

#compute EP band
EP_log_band <- lapply(times_weeks, function(t) {

  S <- get_survival(t)

  if (S$Survival == 0) {
    return(data.frame(
      time = t,
      Survival = 0,
      lower = 0,
      upper = 0
    ))
  }

  theta <- exp(
    k_alpha * (1+N*S$SE) / (N^(1/2)*log(S$Survival))
  )

  lower <- S$Survival^(1/theta)
  upper <- S$Survival^(theta)

  data.frame(
    time = t,
    Survival = S$Survival,
    lower = lower,
    upper = upper
  )
})

EP_log_band <- do.call(rbind, EP_log_band)
EP_log_band
```

I think since question was designed thinking time data was in months and not weeks, this led to the fact $k_\alpha$\[0.23, 0.32\] doesn't exist for our chosen time limits. The code above is what would work if this were a valid interval for our data, I set k = 1 as a placeholder.

**h) Estimate the mean survival time restricted to 400 months. Provide 95% CI for restricted mean survival time.**

400 months likely refers to 400 weeks, so will proceed as if 400 months = 400 weeks.

$\hat\mu(\tau) = \int_0^\tau \hat{S}(t)\,dt$

$\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
=
\sum_{i=1}^{D}
\left(
\int_{t_i}^{\tau} \hat{S}(t)\,dt
\right)^2
\frac{d_i}{Y_i (Y_i - d_i)}$

$\hat{\mu}(\tau)
\pm
z_{1-\alpha/2}
\sqrt{
\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
}$

```{r}
#will use survival package
tau <- 400

fit <- survfit(
  Surv(time, delta) ~ 1,
  data = aneuploid
)

summary(fit, rmean = tau)$table

rmst <- 146.6
SE_rmst <- 27.7

CI_lower <- rmst - 1.96*SE_rmst
CI_upper <- rmst + 1.96*SE_rmst
c(CI_lower, CI_upper)
```

Mean survival time: 146.6

95% CI: \[92, 200\]

**i) Estimate median time to death and find 95% CI for median survival time based on linear CI.**

$\hat{t}_{0.5} = \inf \{ t : \hat{S}(t) \le 0.5 \}$, the first time KM curve drops to 0.5 or below.

$\hat{S}(t) \pm z_{1-\alpha/2}\, SE(\hat{S}(t))$

```{r}
median_idx <- which(km_table$S_t <= 0.5)[1]

median_time <- km_table$time[median_idx]

median_time

z <- 1.96

lower_surv <- km_table$S_t - z * km_table$SE_S_t
upper_surv <- km_table$S_t + z * km_table$SE_S_t
lower_idx <- which(lower_surv <= 0.5)[1]
lower_median <- km_table$time[lower_idx]
upper_idx <- which(upper_surv <= 0.5)[1]
upper_median <- km_table$time[upper_idx]

c(
  Median = median_time,
  Lower95 = lower_median,
  Upper95 = upper_median
)
```

## Question 4.2

25 patients with inoperative lung cancer, all patients followed until death. Start = Nov 1 1979, end = Dec 23 1979. One interim analysis on March 31 1980, when only 13 patients had died.

**a) Estimate the survival function based on the available sample information at the time of interim analysis on 3/31/80. Provide the SE of your estimate.**

time = Days to death
death = Death indicator (1=dead), complete follow-up on all patients
time2 = Days to 3/31/80 or death (interim analysis)
death2 = Death indicator as of 3/31/80 (1=dead, 0=alive)

Time is in days:

$\hat{S}(t)$ = 1 if t less than $t_1$ of interest.

$\hat{S}(t)=∏_{ti≤t}(1−d_i/Y_i)$ if t beyond $t_1$ of interest.

```{r}
rm(list = ls())
data(lung)

N <- nrow(lung)

#order event times
event_times <- sort(unique(lung$time2[lung$death2 == 1])) #on 3/31/80

#start constructing table
km_table <- data.frame(
  time = event_times
)

#compute di, number of deaths at time ti
km_table$d_i <- sapply(
  km_table$time,
  function(t) sum(lung$time2 == t & lung$death2 == 1)
)

#compute ni, number at risk at time ti
km_table$n_i <- sapply(
  km_table$time,
  function(t) sum(lung$time2 >= t)) #observations before t

#survival step terms
km_table$surv_step <- 1 - km_table$d_i / km_table$n_i

#multiply to get estimator for S(t) at each t
km_table$S_t <- cumprod(km_table$surv_step)

#compute sum term 
km_table$greenwood_term <-
  km_table$d_i / (km_table$n_i * (km_table$n_i - km_table$d_i))

#compute cumulative sum
km_table$cum_greenwood <-
  cumsum(km_table$greenwood_term)

#compute greenwood variance by mult cumsum with S(t) estimate for each t
km_table$var_S_t <-
  (km_table$S_t^2) * km_table$cum_greenwood

#compute SE
km_table$SE_S_t <-
  sqrt(km_table$var_S_t)

get_survival <- function(target_time) {

  eligible <- which(km_table$time <= target_time)

  if (length(eligible) == 0) {

    return(data.frame(
      time = target_time,
      Survival = 1,
      SE = 0
    ))

  }

  idx <- max(eligible)

  data.frame(
    time = target_time,
    Survival = km_table$S_t[idx],
    SE = km_table$SE_S_t[idx]
  )
}
```

Here is estimate of survival function as well as SE:

```{r}
km_table %>% select(time, d_i, n_i, S_t, SE_S_t)
```

**b) Use the Brown, Hollandar and Kowar technique to complete right hand tail of the product-limit estimate found in part a).**

Non parametric estimation beyond $t_{max}$ correspond to the 2 most extreme scenarios: complete the tail by an exponential curve picked to give the same value of S($t_{max}$).

$\hat{S}(t)$ for t \> $t_{max}$ is = $exp(t ln[\hat{S}(t_{max})]/t_{max}]$  - plug in t over max t of interest

For t \> 363 (days to death):

```{r}
t_max <- 320 #max time for status check, not entire study

exp(363 * log(get_survival(t_max)$Survival)/t_max) 
log(get_survival(t_max)$Survival)/t_max #-0.004
```

Estimate for t \> 363 days to death (max t for those who died at status check) = 0.21. We can estimate this for any t using exp(t \* -0.004).

**c) Compute the estimate of the survival function and an estimate of its SE using complete follow-up on each patient. Compare this estimate to part a).**

this time around use entire dataset, so time not time2, and death not death2

```{r}
rm(list = ls())
data(lung)

N <- nrow(lung)

#order event times
event_times <- sort(unique(lung$time[lung$death == 1])) #on 3/31/80

#start constructing table
km_table <- data.frame(
  time = event_times
)

#compute di, number of deaths at time ti
km_table$d_i <- sapply(
  km_table$time,
  function(t) sum(lung$time == t & lung$death == 1)
)

#compute ni, number at risk at time ti
km_table$n_i <- sapply(
  km_table$time,
  function(t) sum(lung$time >= t)) #observations before t

#survival step terms
km_table$surv_step <- 1 - km_table$d_i / km_table$n_i

#multiply to get estimator for S(t) at each t
km_table$S_t <- cumprod(km_table$surv_step)

#compute sum term 
km_table$greenwood_term <-
  km_table$d_i / (km_table$n_i * (km_table$n_i - km_table$d_i))

#compute cumulative sum
km_table$cum_greenwood <-
  cumsum(km_table$greenwood_term)

#compute greenwood variance by mult cumsum with S(t) estimate for each t
km_table$var_S_t <-
  (km_table$S_t^2) * km_table$cum_greenwood

#compute SE
km_table$SE_S_t <-
  sqrt(km_table$var_S_t)

get_survival <- function(target_time) {

  eligible <- which(km_table$time <= target_time)

  if (length(eligible) == 0) {

    return(data.frame(
      time = target_time,
      Survival = 1,
      SE = 0
    ))

  }

  idx <- max(eligible)

  data.frame(
    time = target_time,
    Survival = km_table$S_t[idx],
    SE = km_table$SE_S_t[idx]
  )
}
```

Here is estimate of survival function as well as SE

```{r}
km_table %>% select(time, d_i, n_i, S_t, SE_S_t)
```

Relative to part a), this estimate for the survival function seems to lag behind the estimates in a), by a little amount each time such that by the end the discrepancy is just notable.

**d) Estimate the mean time to death restricted to 683 days based on product-limit estimator found in part c)**

$\hat\mu(\tau) = \int_0^\tau \hat{S}(t)\,dt$

$\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
=
\sum_{i=1}^{D}
\left(
\int_{t_i}^{\tau} \hat{S}(t)\,dt
\right)^2
\frac{d_i}{Y_i (Y_i - d_i)}$

$\hat{\mu}(\tau)
\pm
z_{1-\alpha/2}
\sqrt{
\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
}$

```{r}
tau <- 683

#will use survival package
fit <- survfit(
  Surv(time, death) ~ 1,
  data = lung
)

summary(fit, rmean = tau)$table

rmst <- 275.96
SE_rmst <- 39.7343

CI_lower <- rmst - 1.96*SE_rmst
CI_upper <- rmst + 1.96*SE_rmst
c(CI_lower, CI_upper)
```

Mean estimate: 275.96

CI: \[198, 354\]

**e) Estimate the mean time to death by finding the area under the survival curve found in part c. Find SE of estimate.**

$\hat\mu = \int_0^∞ \hat{S}(t)\,dt$ is area under curve estimate

```{r}
tab <- summary(fit)$table

# mean time to death (area under KM curve)
mean_surv <- tab["rmean"]

# standard error of mean survival
SE_mean_surv <- tab["se(rmean)"]

c(mean_survival = mean_surv,
  SE = SE_mean_surv)
```

Mean = 278, SE = 40.4

This estimate takes censoring data into account, however followup is complete so doesn't affect this case.

**f) Compute the usual estimate of the time to death based on complete follow-up data by finding the arithmetic mean of the complete follow-up data. Find SE of this estimate in the usual way as sample SD/sqrt(n). Compare to e)**

SE = $s/sqrt(n)$

```{r}
#arithmetic mean time to death
mean_complete <- mean(lung$time)

#sample size
N <- nrow(lung)

#sample SD
sd_complete <- sd(lung$time)

#standard error
SE_complete <- sd_complete / sqrt(N)

c(mean_complete = mean_complete,
  SE = SE_complete)
```

This estimate does not take censored data into account, however follow-up is complete in this example so this doesn't matter, the estimates are the same as e). The only difference is the SE due to the different variance calculation methods.

## Question 4.4

Times to first exit site infections in months of patients with renal insufficiency was reported.

43 patients had a surgically placed catheter - group1

76 patients had percutaneous placement of catheter - group0

type: group0 = type 2, group1 = type1

delta: infection yes/no

time: times to infection in months

```{r}
rm(list = ls())
data(kidney)
```

**a) For each group plot the estimated survival function. Which technique seems better in delaying the time to infection?**

```{r}
#use survival package
surv_kidney <- Surv(time = kidney$time, event = kidney$delta)

fit_kidney <- survfit(surv_kidney ~ type, data=kidney)

plot(fit_kidney, ylab="Survival Probability", 
     xlab="Months", col = c("red", "blue"))
legend("bottomleft", 
       legend = c("Surgical", "Percutaneous"),
       col = c("red", "blue"),
       lty = 1)
```

The percutaneous placement of the catherer seems to be the method that better delays infection, with advantage most clearly seen \>10months (despite initial disadvantage \<5 months, long term this method fares better).

**b) Estimate the cumulative hazard rate for each group of patients. Provide a crude estimate of the hazard rate at 5 months after placement of catheter in each group.**

We know the derivative of the cumulative hazard = hazard rate.

Look at slope of $\hat H(t)$ at t = 5 months to estimate hazard rate crudely -\> read around 4.5 and 5.5 months, divide by difference of (1)..

```{r}
#plot since already computed
plot(fit_kidney,
     fun = "cumhaz",
     xlab = "Months",
     ylab = "Cumulative Hazard",
     col = c("red", "blue"))
legend("topleft",
       legend = c("Surgical", "Percutaneous"),
       col = c("red", "blue"),
       lty = 1)
#this uses Nelson-Aalen estimate

#crude estimate from slope
summary_5 <- summary(fit_kidney, times = c(4.5, 5.5))

# crude hazard ≈ change in cumulative hazard / change in time
h_5_group1 <- diff(summary_5$cumhaz[summary_5$strata == "type=1"]) / 1
h_5_group0 <- diff(summary_5$cumhaz[summary_5$strata == "type=0"]) / 1
h_5_group1
h_5_group0
```

The crude estimate for both groups is either 0 or very close to 0, from which we can infer no matter which method is used, 5 months after placement the hazard rate is not increasing.

**c) Find a 95% CI for the mean time to first exit site infection restricted to 36 months for both groups.**

$\hat\mu(\tau) = \int_0^\tau \hat{S}(t)\,dt$

$\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
=
\sum_{i=1}^{D}
\left(
\int_{t_i}^{\tau} \hat{S}(t)\,dt
\right)^2
\frac{d_i}{Y_i (Y_i - d_i)}$

$\hat{\mu}(\tau)
\pm
z_{1-\alpha/2}
\sqrt{
\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
}$

```{r}
kidneyfit_36 <- survfit(
  Surv(time, delta) ~ type,
  data = kidney)


tab <- summary(kidneyfit_36, rmean = 36)$table

mean2 <- tab["type=2", "rmean"]
se2   <- tab["type=2", "se(rmean)"]

mean1 <- tab["type=1", "rmean"]
se1   <- tab["type=1", "se(rmean)"]

CI2 <- c(mean2 - 1.96 * se2,
         mean2 + 1.96 * se2)

CI1 <- c(mean1 - 1.96 * se1,
         mean1 + 1.96 * se1)

c(CI2, mean2)
c(CI1, mean1)
```

CI for percutaneous catheter placement: \[25.8, 33.3\], mean time to infection = 29.5

CI for surgically placed catheter: \[15.3, 24.9\] = 20.1

## Question 4.7

## Question 4.8 

Patients admitted to the hospital at a random age and followed until death or end of study -\> there is left truncation and right censoring.

X = patient's age at death

To use the same equations whilst considering left truncation, need to know age at study entrance $L_J$, time of death/censoring $T_j$, #individuals with event at time i $d_i$ and #individuals at risk of experiencing event at time i $Y_i$.

We consider patients at risk if they have already been admitted (age ≤ t), and they have not yet died or been censored (age ≥ t).

Dataset: 26 psychiatric inpatients, on-study time, age at admission, gender

```{r}
rm(list = ls())
data(psych)
#age at admission (age), time of death/censoring (time)
#death: 0 if censored, 1 otherwise
```

**a) Plot the number at risk,** $Y_i$, **as a function of age**

```{r}
#event times
event_age <- psych$age + psych$time

#ordered ages
ages <- sort(unique(event_age))

#calculate Yi
Y <- sapply(ages, function(t) {
  sum(psych$age <= t & event_age >= t)
})

plot(
  ages, Y,
  type = "s",
  xlab = "Age",
  ylab = "Number at risk (Yi)",
  main = "Number at Risk as Function of Age"
)
```

From the plot we can determine that the #individuals at risk decrease as age decreases, there are much more "younger" people at risk. Number at risk decreases as patients die or are censored, so this curve makes sense, as people get older their on study-time ends sooner.

**b) Estimate the conditional survival function for a psychiatric patient who has survived to age 30 without entering a psychiatric hospital.**

$S(t∣T≥30)=\frac {S(30)}{S(t)​}, t≥30$ , using conditional probability definition to consider left truncation.

```{r}

#survival function takes left censoring into account like this
fit_30 <- survfit(
  Surv(pmax(psych$age, 30), event_age, psych$death) ~ 1
)

plot(
  fit_30,
  xlab = "Age",
  ylab = "Conditional survival probability",
  main = "Survival Function Given Survival to Age 30"
)
```
