---
title: "Assignment 3"
format: html
editor: visual
author: "Sofia Quijada"
---

```{r}
library(KMsurv)
library(tidyverse)
library(survival)
```

## Question 4.1

**Effect of ploidy on survival of patients with cancer of the tongue. Using data on ANEUPLOID TUMORS only.**

**a) Estimate survival function at 12 months and 60 months after transplant using table 1.6 data. Get SE of estimates**

type: Tumor DNA profile (1=Aneuploid Tumor, 2=Diploid Tumor) time: Time to death or on-study time, weeks delta: Death indicator (0=alive, 1=dead). 0 = censored in this dataset, as per table 1.6 description.

```{r}
data(tongue)

aneuploid <- subset(tongue, type == 1)

#number of people at risk at beginning of study = number of patients we have data for
N <- nrow(aneuploid)
```

Recall our time is in weeks, change at the end.

$\hat{S}(t)$ = 1 if t less than $t_1$ of interest.

$\hat{S}(t)=∏_{ti≤t}(1−d_i/Y_i)$ if t beyond $t_1$ of interest.

```{r}
#order event times, where delta = 1
event_times <- sort(unique(aneuploid$time[aneuploid$delta == 1]))

#start constructing table
km_table <- data.frame(
  time = event_times
)

#compute di, number of deaths at time ti
km_table$d_i <- sapply(
  km_table$time,
  function(t) sum(aneuploid$time == t & aneuploid$delta == 1) #obs with event (delta = 1) at time t
)

#compute ni, number at risk at time ti
km_table$n_i <- sapply(
  km_table$time,
  function(t) sum(aneuploid$time >= t) #obs before time t
) 
```

Compute $\hat{S}(t)=∏_{ti≤t}(1−d_i/n_i)$ for each row. Multiply them to get $\hat{S}(t)$

```{r}
#survival step terms
km_table$surv_step <- 1 - km_table$d_i / km_table$n_i

#multiply to get estimator for S(t) at each t
km_table$S_t <- cumprod(km_table$surv_step)
```

Compute variance and SE for each estimate.

$\widehat{\mathrm{Var}}(\hat{S}(t)) = \hat{S}(t)^2 \sum_{t_i \le t} \frac{d_i}{Y_i (Y_i - d_i)}$

```{r}
#compute sum term 
km_table$greenwood_term <-
  km_table$d_i / (km_table$n_i * (km_table$n_i - km_table$d_i))

#compute cumulative sum
km_table$cum_greenwood <-
  cumsum(km_table$greenwood_term)

#compute greenwood variance by mult cumsum with S(t) estimate for each t
km_table$var_S_t <-
  (km_table$S_t^2) * km_table$cum_greenwood

#compute SE
km_table$SE_S_t <-
  sqrt(km_table$var_S_t)
```

KM estimator uses the last event time ≤ target time.

```{r}
get_survival <- function(target_time) {

  eligible <- which(km_table$time <= target_time)

  if (length(eligible) == 0) {

    return(data.frame(
      time = target_time,
      Survival = 1,
      SE = 0
    ))

  }

  idx <- max(eligible)

  data.frame(
    time = target_time,
    Survival = km_table$S_t[idx],
    SE = km_table$SE_S_t[idx]
  )
}
#conversion factor from months to weeks is 4.345
#input should be in weeks, 12 or 60 months in weeks
S_12 <- get_survival(12*4.345)
S_60 <- get_survival(60*4.345)

S_12
S_60
```

S(t = 12 months) = 0.65, SE = 0.07

S(t = 60 months) = 0.23, SE = 0.09

**b) Estimate the cumulative hazard rate H(t) at 60 months. Find the standard error of** $\hat{H}(t)$ **Estimate S(60) by exp(-** $\hat{H}(t)$**) and compare to part a)**

$\hat{H}(t) = -ln[\hat{S}(t)]$ from KM estimator

```{r}
H_60 <- -log(S_60$Survival)
H_60
```

Use Nelson Aalen cumulative hazard rate estimator instead.

$\hat{H}(t) = \sum_{t_i \le t} \frac{d_i}{n_i}$, or 0 if $t_1$ of interest bigger than t.

$\mathrm{Var}\!\left[\hat{H}(t)\right] = \sum_{t_i \le t} \frac{d_i}{Y_i^2}$

```{r}
#add nelson-aalen increments
km_table$haz_increment <- km_table$d_i / km_table$n_i

#compute cumulative hazard
km_table$H_t <- cumsum(km_table$haz_increment)

#calculate variance and SE
km_table$var_increment <- km_table$d_i / (km_table$n_i^2)

km_table$var_H_t <- cumsum(km_table$var_increment)

km_table$SE_H_t <- sqrt(km_table$var_H_t)

#estimate at 60 months
idx60 <- max(which(km_table$time <= 60*4.345))
H60 <- km_table$H_t[idx60]
SE_H60 <- km_table$SE_H_t[idx60]
H60
SE_H60
```

$\hat{H}(t = 60)$ = 1.39, SE = 0.375.

Estimate S(60) from H: $\hat{S}(t) = \exp\!\left[-\hat{H}(t)\right]$, compare to S(60)

```{r}
S60_from_NA <- exp(-H60)

S_60$Survival #KM
S60_from_NA 
```

The Nelson-Aalen survival estimator derived from the cumulative hazard rate estimator has a slightly higher value.

**c) Find a 95% linear CI for S(60)**

$\hat{S}(t) \pm z_{0.975}\,\mathrm{SE}\!\left(\hat{S}(t)\right)$

```{r}
z <- 1.96

lower_linear <- S_60$Survival - z * S_60$SE
upper_linear <- S_60$Survival + z * S_60$SE

c(lower_linear, upper_linear)
```

\[0.042, 0.42\]. This is a big confidence interval.

**d) Find a 95% log-transformed CI for S(60)**

Confidence interval =$[\hat{S}(t_1)^{1/\theta}, \hat{S}(t_1)^{ \theta}]$,

Where $\theta = exp({z_{0.975}SE(S(t_1))/ln(S(t_1)})$

```{r}
theta <- exp(z * S_60$SE / log(S_60$Survival))
lower_log <- S_60$Survival^(1/theta)
upper_log <- S_60$Survival^theta
c(lower_log, upper_log)
```

\[0.187, 0.273\] 95% CI for S(60) = 0.23, much more reasonable

**e) Find a 95% arcsine-square root CI for S(60)**

Not computing this manually. Use "survival" package.

```{r}
fit <- survfit(Surv(time, delta) ~ 1, data = aneuploid, conf.type = "arcsin")
summary(fit, times = c(12*4.345, 60*4.345))
```

\[0.0744, 0.436\] is 95% arcsin square root survival CI.

**f) Using the log transformation find a 95% EP confidence band for the survival function over the range 3-6 yrs (36-72 months)**

range: 36 \< t \< 72 months

n = sample size

$a_L = \frac{n Var(t_L),}{1+n Var(t_L))}$ , $a_U = \frac{n Var(t_U),}{1+n Var(t_U))}$

Confidence band =$[\hat{S}(t_1)^{1/\theta}, \hat{S}(t_1)^{ \theta}]$,

Where $\theta =exp(\frac{c_\alpha(a_L, a_U) SE(S(t)))}{ln[\hat{S}(t)]})$

$c_\alpha$ found from table C.3 of textbook using values of $a_L$ and $a_u$

```{r}
# bounds in weeks
lower_t <- 36 * 4.345
upper_t <- 72 * 4.345

lower_S <- get_survival(lower_t)
upper_S <- get_survival(upper_t)

# variance = SE^2
a_L <- N * (lower_S$SE)^2 / (1 + N * (lower_S$SE)^2) #0.234
a_U <- N * (upper_S$SE)^2 / (1 + N * (upper_S$SE)^2) #0.321

# from Table C.3 - between 3.06 and 3.02 
c_alpha <- 3.04

theta_lower <- exp(
  c_alpha * lower_S$SE / log(lower_S$Survival)
)

theta_upper <- exp(
  c_alpha * upper_S$SE / log(upper_S$Survival)
)

# correct ordering
lower_band_36 <- lower_S$Survival^(1/theta_lower)
upper_band_36 <- lower_S$Survival^(theta_lower)

lower_band_72 <- upper_S$Survival^(1/theta_upper)
upper_band_72 <- upper_S$Survival^(theta_upper)

print(c(lower_band_36, upper_band_36))

print(c(lower_band_72, upper_band_72))
```

This is just for the end points, here is for the entire time period:

```{r}
# bounds in weeks
lower_t <- 36 * 4.345
upper_t <- 72 * 4.345

lower_S <- get_survival(lower_t)
upper_S <- get_survival(upper_t)

# variance = SE^2
a_L <- N * (lower_S$SE)^2 / (1 + N * (lower_S$SE)^2) #0.234
a_U <- N * (upper_S$SE)^2 / (1 + N * (upper_S$SE)^2) #0.321

# from Table C.3 - between 3.06 and 3.02 
c_alpha <- 3.04

times_weeks <- km_table$time[
  km_table$time >= lower_t &
  km_table$time <= upper_t
]

#compute EP band
EP_band <- lapply(times_weeks, function(t) {

  S <- get_survival(t)

  if (S$Survival == 0) {
    return(data.frame(
      time = t,
      Survival = 0,
      lower = 0,
      upper = 0
    ))
  }

  theta <- exp(
    c_alpha * S$SE / log(S$Survival)
  )

  lower <- S$Survival^(1/theta)
  upper <- S$Survival^(theta)

  data.frame(
    time = t,
    Survival = S$Survival,
    lower = lower,
    upper = upper
  )
})

EP_band <- do.call(rbind, EP_band)
EP_band
```

Since data given is in weeks, there are little observations that go beyond 36 months. I suspect the questions were designed thinking the "time" data was in weeks.

**g) Using log transformation find 95% Hall-Wellner confidence band for the survival function over the range 3-6 yrs (36-72 months)**

$a_L = \frac{n Var(t_L),}{1+n Var(t_L))}$ , $a_U = \frac{n Var(t_U),}{1+n Var(t_U))}$

Confidence band =$[\hat{S}(t_1)^{1/\theta}, \hat{S}(t_1)^{ \theta}]$,

Where $\theta =exp(\frac{k_\alpha(a_L, a_U)(1 + nVar(S(t))}{n^{1/2}ln[\hat{S}(t)]})$

$k_\alpha$ from table C.4 of Appendix C.

```{r}
# bounds in weeks
lower_t <- 36 * 4.345
upper_t <- 72 * 4.345

lower_S <- get_survival(lower_t)
upper_S <- get_survival(upper_t)

# variance = SE^2
a_L <- N * (lower_S$SE)^2 / (1 + N * (lower_S$SE)^2) #0.234
a_U <- N * (upper_S$SE)^2 / (1 + N * (upper_S$SE)^2) #0.321

# from Table C.3 - between 3.06 and 3.02 
k_alpha <- 1 #doesn't seem to exist :c, put in 1 as a placeholder

times_weeks <- km_table$time[
  km_table$time >= lower_t &
  km_table$time <= upper_t
]

#compute EP band
EP_log_band <- lapply(times_weeks, function(t) {

  S <- get_survival(t)

  if (S$Survival == 0) {
    return(data.frame(
      time = t,
      Survival = 0,
      lower = 0,
      upper = 0
    ))
  }

  theta <- exp(
    k_alpha * (1+N*S$SE) / (N^(1/2)*log(S$Survival))
  )

  lower <- S$Survival^(1/theta)
  upper <- S$Survival^(theta)

  data.frame(
    time = t,
    Survival = S$Survival,
    lower = lower,
    upper = upper
  )
})

EP_log_band <- do.call(rbind, EP_log_band)
EP_log_band
```

I think since question was designed thinking time data was in months and not weeks, this led to the fact $k_\alpha$\[0.23, 0.32\] doesn't exist for our chosen time limits. The code above is what would work if this were a valid interval for our data, I set k = 1 as a placeholder.

**h) Estimate the mean survival time restricted to 400 months. Provide 95% CI for restricted mean survival time.**

400 months likely refers to 400 weeks, so will proceed as if 400 months = 400 weeks.

$\hat\mu(\tau) = \int_0^\tau \hat{S}(t)\,dt$

$\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
=
\sum_{i=1}^{D}
\left(
\int_{t_i}^{\tau} \hat{S}(t)\,dt
\right)^2
\frac{d_i}{Y_i (Y_i - d_i)}$

$\hat{\mu}(\tau)
\pm
z_{1-\alpha/2}
\sqrt{
\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
}$

```{r}
#will use survival package
tau <- 400

fit <- survfit(
  Surv(time, delta) ~ 1,
  data = aneuploid
)

summary(fit, rmean = tau)$table

rmst <- 146.6
SE_rmst <- 27.7

CI_lower <- rmst - 1.96*SE_rmst
CI_upper <- rmst + 1.96*SE_rmst
c(CI_lower, CI_upper)
```

Mean survival time: 146.6

95% CI: \[92, 200\]

**i) Estimate median time to death and find 95% CI for median survival time based on linear CI.**

$\hat{t}_{0.5} = \inf \{ t : \hat{S}(t) \le 0.5 \}$, the first time KM curve drops to 0.5 or below.

$\hat{S}(t) \pm z_{1-\alpha/2}\, SE(\hat{S}(t))$

```{r}
median_idx <- which(km_table$S_t <= 0.5)[1]

median_time <- km_table$time[median_idx]

median_time

z <- 1.96

lower_surv <- km_table$S_t - z * km_table$SE_S_t
upper_surv <- km_table$S_t + z * km_table$SE_S_t
lower_idx <- which(lower_surv <= 0.5)[1]
lower_median <- km_table$time[lower_idx]
upper_idx <- which(upper_surv <= 0.5)[1]
upper_median <- km_table$time[upper_idx]

c(
  Median = median_time,
  Lower95 = lower_median,
  Upper95 = upper_median
)
```

## Question 4.2

25 patients with inoperative lung cancer, all patients followed until death. Start = Nov 1 1979, end = Dec 23 1979. One interim analysis on March 31 1980, when only 13 patients had died.

**a) Estimate the survival function based on the available sample information at the time of interim analysis on 3/31/80. Provide the SE of your estimate.**

time = Days to death
death = Death indicator (1=dead), complete follow-up on all patients
time2 = Days to 3/31/80 or death (interim analysis)
death2 = Death indicator as of 3/31/80 (1=dead, 0=alive)

Time is in days:

$\hat{S}(t)$ = 1 if t less than $t_1$ of interest.

$\hat{S}(t)=∏_{ti≤t}(1−d_i/Y_i)$ if t beyond $t_1$ of interest.

```{r}
rm(list = ls())
data(lung)

N <- nrow(lung)

#order event times
event_times <- sort(unique(lung$time2[lung$death2 == 1])) #on 3/31/80

#start constructing table
km_table <- data.frame(
  time = event_times
)

#compute di, number of deaths at time ti
km_table$d_i <- sapply(
  km_table$time,
  function(t) sum(lung$time2 == t & lung$death2 == 1)
)

#compute ni, number at risk at time ti
km_table$n_i <- sapply(
  km_table$time,
  function(t) sum(lung$time2 >= t)) #observations before t

#survival step terms
km_table$surv_step <- 1 - km_table$d_i / km_table$n_i

#multiply to get estimator for S(t) at each t
km_table$S_t <- cumprod(km_table$surv_step)

#compute sum term 
km_table$greenwood_term <-
  km_table$d_i / (km_table$n_i * (km_table$n_i - km_table$d_i))

#compute cumulative sum
km_table$cum_greenwood <-
  cumsum(km_table$greenwood_term)

#compute greenwood variance by mult cumsum with S(t) estimate for each t
km_table$var_S_t <-
  (km_table$S_t^2) * km_table$cum_greenwood

#compute SE
km_table$SE_S_t <-
  sqrt(km_table$var_S_t)

get_survival <- function(target_time) {

  eligible <- which(km_table$time <= target_time)

  if (length(eligible) == 0) {

    return(data.frame(
      time = target_time,
      Survival = 1,
      SE = 0
    ))

  }

  idx <- max(eligible)

  data.frame(
    time = target_time,
    Survival = km_table$S_t[idx],
    SE = km_table$SE_S_t[idx]
  )
}
```

Here is estimate of survival function as well as SE:

```{r}
km_table %>% select(time, d_i, n_i, S_t, SE_S_t)
```

**b) Use the Brown, Hollandar and Kowar technique to complete right hand tail of the product-limit estimate found in part a).**

Non parametric estimation beyond $t_{max}$ correspond to the 2 most extreme scenarios: complete the tail by an exponential curve picked to give the same value of S($t_{max}$).

$\hat{S}(t)$ for t \> $t_{max}$ is = $exp(t ln[\hat{S}(t_{max})]/t_{max}]$  - plug in t over max t of interest

For t \> 363 (days to death):

```{r}
t_max <- 320 #max time for status check, not entire study

exp(363 * log(get_survival(t_max)$Survival)/t_max) 
log(get_survival(t_max)$Survival)/t_max #-0.004
```

Estimate for t \> 363 days to death (max t for those who died at status check) = 0.21. We can estimate this for any t using exp(t \* -0.004).

**c) Compute the estimate of the survival function and an estimate of its SE using complete follow-up on each patient. Compare this estimate to part a).**

this time around use entire dataset, so time not time2, and death not death2

```{r}
rm(list = ls())
data(lung)

N <- nrow(lung)

#order event times
event_times <- sort(unique(lung$time[lung$death == 1])) #on 3/31/80

#start constructing table
km_table <- data.frame(
  time = event_times
)

#compute di, number of deaths at time ti
km_table$d_i <- sapply(
  km_table$time,
  function(t) sum(lung$time == t & lung$death == 1)
)

#compute ni, number at risk at time ti
km_table$n_i <- sapply(
  km_table$time,
  function(t) sum(lung$time >= t)) #observations before t

#survival step terms
km_table$surv_step <- 1 - km_table$d_i / km_table$n_i

#multiply to get estimator for S(t) at each t
km_table$S_t <- cumprod(km_table$surv_step)

#compute sum term 
km_table$greenwood_term <-
  km_table$d_i / (km_table$n_i * (km_table$n_i - km_table$d_i))

#compute cumulative sum
km_table$cum_greenwood <-
  cumsum(km_table$greenwood_term)

#compute greenwood variance by mult cumsum with S(t) estimate for each t
km_table$var_S_t <-
  (km_table$S_t^2) * km_table$cum_greenwood

#compute SE
km_table$SE_S_t <-
  sqrt(km_table$var_S_t)

get_survival <- function(target_time) {

  eligible <- which(km_table$time <= target_time)

  if (length(eligible) == 0) {

    return(data.frame(
      time = target_time,
      Survival = 1,
      SE = 0
    ))

  }

  idx <- max(eligible)

  data.frame(
    time = target_time,
    Survival = km_table$S_t[idx],
    SE = km_table$SE_S_t[idx]
  )
}
```

Here is estimate of survival function as well as SE

```{r}
km_table %>% select(time, d_i, n_i, S_t, SE_S_t)
```

Relative to part a), this estimate for the survival function seems to lag behind the estimates in a), by a little amount each time such that by the end the discrepancy is just notable.

**d) Estimate the mean time to death restricted to 683 days based on product-limit estimator found in part c)**

$\hat\mu(\tau) = \int_0^\tau \hat{S}(t)\,dt$

$\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
=
\sum_{i=1}^{D}
\left(
\int_{t_i}^{\tau} \hat{S}(t)\,dt
\right)^2
\frac{d_i}{Y_i (Y_i - d_i)}$

$\hat{\mu}(\tau)
\pm
z_{1-\alpha/2}
\sqrt{
\widehat{\mathrm{Var}}\!\left[\hat{\mu}(\tau)\right]
}$

```{r}
tau <- 683

#will use survival package
fit <- survfit(
  Surv(time, death) ~ 1,
  data = lung
)

summary(fit, rmean = tau)$table

rmst <- 275.96
SE_rmst <- 39.7343

CI_lower <- rmst - 1.96*SE_rmst
CI_upper <- rmst + 1.96*SE_rmst
c(CI_lower, CI_upper)
```

Mean estimate: 275.96

CI: \[198, 354\]

**e) Estimate the mean time to death by finding the area under the survival curve found in part c. Find SE of estimate.**

$\hat\mu = \int_0^∞ \hat{S}(t)\,dt$ is area under curve estimate

```{r}
tab <- summary(fit)$table

# mean time to death (area under KM curve)
mean_surv <- tab["rmean"]

# standard error of mean survival
SE_mean_surv <- tab["se(rmean)"]

c(mean_survival = mean_surv,
  SE = SE_mean_surv)
```

Mean = 278, SE = 40.4

This estimate takes censoring data into account, however followup is complete so doesn't affect this case.

**f) Compute the usual estimate of the time to death based on complete follow-up data by finding the arithmetic mean of the complete follow-up data. Find SE of this estimate in the usual way as sample SD/sqrt(n). Compare to e)**

SE = $s/sqrt(n)$

```{r}
#arithmetic mean time to death
mean_complete <- mean(lung$time)

#sample size
N <- nrow(lung)

#sample SD
sd_complete <- sd(lung$time)

#standard error
SE_complete <- sd_complete / sqrt(N)

c(mean_complete = mean_complete,
  SE = SE_complete)
```

This estimate does not take censored data into account, however follow-up is complete in this example so this doesn't matter, the estimates are the same as e). The only difference is the SE due to the different variance calculation methods.
